---
title: "Permutation Tests and ANOVA"
author: "Ethan Tsao"
date: "2024-10-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```

# Part 1: Two sample permutation test

## a) Permutation test function

```{r}
perm_test <- function(data, reps){
  
  perm_t <- rep(NA, reps)
  
  names(data) <- c("y", "x")
  
  for(i in 1:reps){
    permdata <- data
    permdata$y <- sample(permdata$y)
    perm_t[i] <- t.test(y ~ x,
                        data=permdata)$statistic
  }

  t0 <- t.test(y ~ x, data=data)$statistic

  p_value <- (sum(perm_t >= abs(t0)) + sum(perm_t <= -abs(t0)))/reps
  
  return(p_value)
}
```

## b) Testing your function

### (i) Table 2.1 on page 23 in our textbook

```{r}
Modified_Mortar <- c(16.85, 16.40, 17.21, 16.35, 16.52, 17.04, 16.96, 17.15, 16.59, 16.57)
Unmodified_Mortar <- c(16.62, 16.75, 17.37, 17.12, 16.98, 16.87, 17.34, 17.02, 17.08, 17.27)

mortar.df <- data.frame(
  y = c(Modified_Mortar, Unmodified_Mortar),
  x = factor(rep(c("Modified", "Unmodified"), each = 10))
)

mortar.df
```

```{r, fig.show="hold", out.width="50%"}
# Boxplot
ggplot(data=mortar.df, aes(x = x, y = y)) +
  geom_boxplot() +
  ggtitle("Boxplot of Modified vs Unmodified Mortar") +
  ylab("Strength") +
  xlab("Group")

# Scatterplot
ggplot(data=mortar.df, aes(x = x, y = y)) +
  geom_point() +
  ggtitle("Scatterplot of Modified vs Unmodified Mortar") +
  ylab("Strength") +
  xlab("Group")
```
```{r}
set.seed(123)

perm_test(mortar.df, reps = 1000)
```

```{r}
t.test(y ~ x, data = mortar.df)
```
### Observations from Side-by-Side Boxplots and Scatter Diagrams:
* From the boxplot, we observe that both groups (Modified and Unmodified Mortar) have fairly similar central tendencies, but the Unmodified Mortar group seems to have slightly higher values overall.

* The scatterplot further confirms that while there is some overlap in the strength values, the Unmodified Mortar group generally has slightly higher measurements than the Modified Mortar group.

### Permutation Test Results:
* The permutation test returned a p-value of 0.049, which is very close to the conventional significance level of 0.05. This suggests that there might be a statistically significant difference between the two groups.

### t-Test Results:
* The t-test resulted in a t-statistic of -2.19 and a p-value of 0.043, which is also statistically significant at the 5% level, indicating that there is likely a difference between the means of the two groups.

* The 95% confidence interval for the difference in means is [-0.546, -0.0098], meaning the difference is small but statistically significant.

### Comparison of Results:
* Both the permutation test and t-test yield similar conclusions: the p-values are below 0.05, suggesting a statistically significant difference between the two groups.

* The slight differences in p-values (0.049 for the permutation test and 0.043 for the t-test) could be due to the inherent differences in how each test works. The permutation test does not make assumptions about normality, whereas the t-test assumes normality and equal variances which might not be entirely true for these data.

### (ii) Non-normal data with a true difference in means

```{r, fig.show="hold", out.width="50%"}
library(readr)
weird_data <- read_csv("/Users/ethantsao/PSTAT 122/weird_data.csv")

ggplot(data=weird_data, aes(x = x, y = y)) +
  geom_boxplot() +
  ggtitle("Boxplot of Weird Data") +
  ylab("y") +
  xlab("Group")

ggplot(data=weird_data, aes(x = x, y = y)) +
  geom_point() +
  ggtitle("Scatterplot of Weird Data") +
  ylab("y") +
  xlab("Group")
```
```{r}
set.seed(123)

perm_test(weird_data, reps = 1000)
```

```{r}
t.test(y ~ x, data = weird_data)
```
### Observations from Side-by-Side Boxplots and Scatter Diagrams:
* The boxplot shows a notable difference between the two groups. Group 1 has higher median values than Group 2, and there is less overlap between the groups compared to the mortar dataset.

* The scatterplot also shows a clearer separation between Group 1 and Group 2, indicating a true difference in distributions.

### Permutation Test Results:
* The permutation test returned a p-value of 0.033, indicating that the difference between the two groups is statistically significant at the 5% level.

### t-Test Results:
* The t-test produced a t-statistic of 2.08 with a p-value of 0.0738, which is not significant at the 5% level but is close.

* The 95% confidence interval for the difference in means is [-0.101, 1.734], suggesting that while the data shows a difference, the t-test is not as confident due to the non-normality of the data.

### Comparison of Results:
* The permutation test yielded a significant result, whereas the t-test did not. This occurred because the t-test relies on assumptions of normality, and since the data is highly non-normal, the t-test underestimates this difference.

* The permutation test, on the other hand, does not rely on these assumptions and is more accurate in this case, hence the significant p-value.

* Since we know that there is a true difference in means, the t-test is likely underpowered which could lead to a Type II error (failing to reject the null hypothesis when it is false).

### (iii) Non-normal data with no true difference in means

```{r, fig.show="hold", out.width="50%"}
library(readr)
weird_data2 <- read_csv("/Users/ethantsao/PSTAT 122/weird_data2.csv")

ggplot(data=weird_data2, aes(x = x, y = y)) +
  geom_boxplot() +
  ggtitle("Boxplot of Weird Data2") +
  ylab("y") +
  xlab("Group")

ggplot(data=weird_data2, aes(x = x, y = y)) +
  geom_point() +
  ggtitle("Scatterplot of Weird Data2") +
  ylab("y") +
  xlab("Group")
```
```{r}
set.seed(123)

perm_test(weird_data2, reps = 1000)
```

```{r}
t.test(y~x, data = weird_data2)
```
### Observations from Side-by-Side Boxplots and Scatter Diagrams:
* Both the boxplot and scatterplot show that the two groups (Group 1 and Group 2) have highly overlapping distributions, with no clear visual distinction between their central tendencies or spreads.
	
### Permutation Test Results:
* The permutation test yielded a p-value of 0.062, which is slightly above the 0.05 threshold, indicating no statistically significant difference between the two groups.

### t-Test Results:
* The t-test resulted in a t-statistic of 2.097 and a p-value of 0.0496, which is just below the 0.05 threshold, suggesting a significant difference between the groups.

### Comparison of Results:
* There is a discrepancy between the two tests: the t-test finds a significant difference, while the permutation test does not.

* Since we know there is no true difference in means, the t-test result might be a false positive, likely leading to a Type I error (incorrectly rejecting the null hypothesis when it is true).

* The permutation test’s p-value being just above 0.05 suggests it is more conservative in this case, which may be preferable given what we know to be true about the data.

# Part 2: ANOVA

## Exercise 3.9

### Part a)

### Null Hypothesis (H0):
The mean tensile strengths of the cement are the same across all four mixing techniques.

H0: $\mu_1$ = $\mu_2$ = $\mu_3$ = $\mu_4$

### Alternative Hypothesis (H1):
At least one of the mixing techniques leads to a different mean tensile strength.

H1: $\text{At least one }$ $\mu_i$ $\text{ is different}$

### Significance Level:
We will use $\alpha = 0.05$  as the significance level.

### Conducting the ANOVA Test:
Using an ANOVA test, we can compare the mean tensile strengths across the four groups. ANOVA is the appropriate test because we are comparing more than two group means.

```{r}
data <- data.frame(
  tensile_strength = c(3129, 3000, 2865, 2890,
                       3200, 3150, 2985, 3050,
                       2800, 2900, 2985, 3000,
                       2600, 2765, 2700, 2600), 
  technique = factor(rep(1:4, each = 4))
)

anova_result <- aov(tensile_strength ~ technique, data = data)
summary(anova_result)
```

### Conclusion:
* Since the p-value is 0.000387, which is much smaller than the significance level $\alpha = 0.05$ , we reject the null hypothesis. This means there is strong evidence that at least one of the mixing techniques significantly affects the tensile strength of the cement.

### Part b)
```{r}
ggplot(data, aes(x = technique, y = tensile_strength)) +
  geom_boxplot() +
  ggtitle("Comparison of Tensile Strengths by Mixing Technique") +
  ylab("Tensile Strength (lb/in^2") +
  xlab("Mixing Technique")
```

### Mixing Techniques 1 and 2:
Both techniques 1 and 2 exhibit relatively higher tensile strength, with technique 2 having the highest median and range. Technique 1 has a slightly lower median but still a large spread.

### Mixing Technique 3:
Technique 3 shows a moderate tensile strength but has a tighter spread compared to techniques 1 and 
Its median is lower than those two techniques.

### Mixing Technique 4:
Technique 4 has the lowest tensile strength among the four techniques. It also shows less variability, with its interquartile range being smaller.

### Conclusions:

* There appears to be a clear difference in the distributions of tensile strengths across the different techniques. Techniques 1 and 2 seem to have stronger tensile results, while Technique 4 is the weakest.

* These graphical results are consistent with the findings from the ANOVA test, where we rejected the null hypothesis, showing that there is a significant difference in the tensile strengths across the four mixing techniques.
